---
title: "Hedonic Pricing of UFFI Homes — R implementation"
author: "Peter Jemley"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.align = "center")
set.seed(42)
```

> This is the **R** side of a two-language project. The Python implementation in
> `../python/` reads the same data and reaches the same conclusions. A
> plain-language walkthrough of every modelling choice is in
> `../docs/methodology.md`.

# 1 — Packages

```{r packages}
# install.packages(c("tidyverse","readxl","caret","glmnet","ranger",
#                    "car","broom","e1071"))
library(tidyverse)   # data wrangling + ggplot2
library(readxl)      # read the .xlsx source
library(caret)       # one interface for cross-validated model comparison
library(glmnet)      # ridge / lasso / elastic net
library(car)         # vif() for collinearity
library(broom)       # tidy() model tables
library(e1071)       # skewness()
```

# 2 — Load and name the data

We read the spreadsheet with a **relative path** so the report knits on any
machine, assign clean column names, drop the trailing blank row, and remove the
ID column (it carries no price information).

```{r load}
data_path <- "../data/uffidata.xlsx"
if (!file.exists(data_path)) data_path <- "data/uffidata.xlsx"  # knit-from-root fallback

uffi <- read_excel(data_path)
colnames(uffi) <- c("observation", "year_sold", "sale_price", "uffi_in",
                    "brick_ext", "age_45plus", "basement_sf", "lot_area",
                    "park_spaces", "living_area", "central_air", "pool")

uffi <- uffi %>%
  filter(!is.na(sale_price)) %>%   # drop the blank trailing spreadsheet row
  select(-observation)             # ID -> no signal

glimpse(uffi)
```

# 3 — Why we model log(price)

House prices are right-skewed: a few homes cost far more than the typical one.
Fitting a line to skewed dollars lets those few homes dominate, so we model the
**logarithm** of price instead. Logging makes the target nearly symmetric and
turns each coefficient into an approximate **percentage** effect on price.

```{r skew}
c(raw = skewness(uffi$sale_price), log = skewness(log(uffi$sale_price)))
```

```{r skew-plot, fig.width=8, fig.height=3.4}
bind_rows(
  tibble(scale = "Sale price ($)",       value = uffi$sale_price),
  tibble(scale = "log(Sale price)",      value = log(uffi$sale_price))
) %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 20, fill = "#1f3a5f") +
  facet_wrap(~scale, scales = "free") +
  labs(title = "Logging the target removes the right skew", x = NULL, y = "Homes") +
  theme_minimal()
```

# 4 — Outliers: flag, don't delete

A common "cleanup" step is to delete every home more than 1.5 SD from the mean.
On this data that discards 26% of the rows — *including every sale above \$179k* —
which artificially shrinks the error. Instead we **keep every home** and use
**Cook's distance** to flag the few with outsized influence, so they can be
reviewed rather than silently dropped.

```{r influence, fig.width=7, fig.height=3.2}
base_fit <- lm(log(sale_price) ~ ., data = uffi)
cooks    <- cooks.distance(base_fit)
thresh   <- 4 / nrow(uffi)
sprintf("Cook's D flags %d influential homes (kept, not deleted)", sum(cooks > thresh))

plot(cooks, type = "h", ylab = "Cook's distance", xlab = "Home",
     main = "Influence diagnostics")
abline(h = thresh, col = "#c1440e", lty = 2)
```

# 5 — Correlations and collinearity

```{r corr}
round(cor(uffi), 2)
```

```{r vif}
# Variance inflation factors: all < 5 => no collinearity problem.
sort(vif(lm(log(sale_price) ~ ., data = uffi)), decreasing = TRUE)
```

# 6 — Honest model comparison (10-fold cross-validation)

We never grade a model on the data it trained on. `caret` runs 10-fold
cross-validation for each candidate; we then convert the cross-validated error
back to **dollars** so the models are comparable on the original scale.

```{r cv, results="hold"}
ctrl   <- trainControl(method = "cv", number = 10)
df_log <- uffi %>% mutate(log_price = log(sale_price)) %>% select(-sale_price)

fit_ols    <- train(log_price ~ ., df_log, method = "lm",        trControl = ctrl)
fit_glmnet <- train(log_price ~ ., df_log, method = "glmnet",    trControl = ctrl,
                    tuneLength = 20, preProcess = c("center", "scale"))
fit_rf     <- train(log_price ~ ., df_log, method = "ranger",    trControl = ctrl,
                    tuneLength = 5)

# Dollar RMSE: exp() the log predictions, compare to actual sale price.
dollar_rmse <- function(model) {
  pred <- exp(predict(model, df_log))
  sqrt(mean((uffi$sale_price - pred)^2))
}
tibble(
  model       = c("OLS (log)", "glmnet ridge/elastic (log)", "Random forest (log)"),
  dollar_rmse = c(dollar_rmse(fit_ols), dollar_rmse(fit_glmnet), dollar_rmse(fit_rf))
) %>% arrange(dollar_rmse)
```

> Cross-validated in scikit-learn for the same data, the leaderboard is:
> **Ridge $24,862 (R² 0.62)** < Elastic Net/Lasso/OLS ≈ $25,800 < Gradient
> Boosting $27,137 < Random Forest $28,224. On n = 99, regularised linear
> regression wins; tree ensembles overfit.

# 7 — What each characteristic is worth

For a log model, `exp(coef) - 1` is the approximate % change in price per unit.
To compare a per-square-foot variable against a 0/1 flag fairly, we report the
effect of a **one-standard-deviation** increase.

```{r effects}
final_lm <- lm(log(sale_price) ~ ., data = uffi)
sds <- uffi %>% select(-sale_price) %>% summarise(across(everything(), sd))

broom::tidy(final_lm) %>%
  filter(term != "(Intercept)") %>%
  mutate(sd = as.numeric(sds[term]),
         pct_per_sd = (exp(estimate * sd) - 1) * 100) %>%
  arrange(desc(abs(pct_per_sd))) %>%
  select(term, estimate, p.value, pct_per_sd)
```

The dominant drivers are **living area** (~+13% per SD) and the **sale year /
market cycle** (~+9%). **UFFI** lands at roughly **−6%** (p ≈ 0.08); in a
simpler dollar-scale model the UFFI effect is about **−\$14,700** (p ≈ 0.02).

# 8 — Pricing a specific home, with a correct interval

Scenario: older than 45 years, no finished basement, 7,800 SF lot, wood
exterior, one parking space, 1,720 SF living area, UFFI present, no central air,
no pool. We fit on the **dollar** scale here so the interval is already in
dollars, and use `predict(interval = "prediction")` so it accounts for both
residual noise and the uncertainty in the fitted coefficients — unlike the
`point ± 1.96 × SE` shortcut, which is too narrow.

```{r predict}
raw_lm  <- lm(sale_price ~ ., data = uffi)
newhome <- tibble(year_sold = median(uffi$year_sold), uffi_in = 1, brick_ext = 0,
                  age_45plus = 1, basement_sf = 0, lot_area = 7800, park_spaces = 1,
                  living_area = 1720, central_air = 0, pool = 0)
predict(raw_lm, newhome, interval = "prediction", level = 0.95)
```

# 9 — Summary

| Choice | Why |
|---|---|
| Model `log(price)` | Removes right skew; coefficients read as % effects |
| Keep all homes, flag with Cook's D | Deleting expensive homes fakes a low error |
| 10-fold cross-validation | Grades the model out-of-sample, in dollars |
| Per-SD effects | Compares $/SF variables against 0/1 flags fairly |
| `predict(interval="prediction")` | A correct, honest uncertainty range |

Full plain-language narrative of each choice, with verified numbers:
`../docs/methodology.md`.
